\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-21 paper.tex
natural evolution strategies

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likehood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem with variational optimization. Effectively, the adversarial
variational optimization procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the respective
hierarchical model matches the observed data.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In this note, we develop a likelihood-free inference algorithm for fitting a
forward non-differentiable generative model to observed data. We propose to
adapt the adversarial training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem with variational
optimization~\citep{2012arXiv1212.4507S}. The adversarial variational
optimization procedure results in learning a proposal distribution over
simulator parameters, hence producing an abitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

\subsection{Variational optimization}

Following \citep{2012arXiv1212.4507S}, variational optimization (VO) (also known as the search gradient algorithm~\citep{2011arXiv1106.4487W} related to evolution strategies) is a general
optimization technique that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize, VO
is based on the simple fact that
\begin{equation}
    \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c}) \leq \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] = U(\psi),
\end{equation}
where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\mathbf{c}$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\mathbf{c}^* = \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c})$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
\begin{align}\label{eqn:vo-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
    &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c})]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log q_\psi(\mathbf{c})$ of the proposal
is known and that one can evaluate $f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then
one can construct empirical estimates of Eqn.~\ref{eqn:vo-grad}, which can
in turn be used to perform stochastic gradient descent (or a variant thereof)
in order to minimize $U(\psi)$.

@@

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known likelihood)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \}_{m=1}^M \sim p_r$, $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$, $\{\epsilon_m \}_{m=1}^M \sim U[0,1]$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$L^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{RMSProp}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M L^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
            \State{$\psi \leftarrow \text{RMSProp}(\tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m) + \gamma  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this work, we propose to perform variational optimization
on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
over the parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align}
respectively over $\phi$ and $\psi$.
When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
obtained by evaluating the exact and known gradient of $U_d$ over
mini-batches of true and generated data, as ordinarily
done in stochastic gradient descent. When updating $g$, gradient estimates
of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
\end{equation}
which we can approximate with mini-batches of generated data
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
\end{equation}
for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
In practice, using
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
\end{equation}
works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
However, this formulation does not necessarily guarantee that the
proposal distribution $q_\psi$ will place its mass arbitrarily tight
around the parameters of interest $\theta^*$. For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] + \gamma H(q_\psi)
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
proposal distributions with larger entropy are not penalized, which may result
in learning a smeared variation of the original simulator.

For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
optimization procedure, as built on top of WGAN-GP.
Let us note however that the variational relaxation could similarly be coupled with
other variants of GANs. \glnote{add note on rmsprop vs. adam}

>>>>

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likehood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variationial upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution  matches the observed data.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In this note, we develop a likelihood-free inference algorithm for fitting a
forward non-differentiable generative model to observed data. We propose to
adapt the adversarial training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper bounds of the adversarial objectives
respectively with stochastic gradient descent and  natural evolution strategies~\citep{2011arXiv1106.4487W}.
The procedure results in learning a proposal distribution over
simulator parameters, hence producing an abitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

\subsection{Natural evolution strategies}

Evolution strategies~\citep{2011arXiv1106.4487W} and variational optimization~\cite{2012arXiv1212.4507S} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c}) \leq \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] = U(\psi),
\end{equation}
where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\mathbf{c}$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\mathbf{c}^* = \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c})$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
    &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c})]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log q_\psi(\mathbf{c})$ of the proposal
is known and that one can evaluate $f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then
one can construct empirical estimates of Eqn.~\ref{eqn:approx-grad}, which can
in turn be used to perform stochastic gradient descent (or a variant thereof)
in order to minimize $U(\psi)$.

In this simple form, stochastic gradient descent based on
Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
of scale invariance. To avoid the dependence on the parameterization of the
proposal distribution, natural gradients~\citep{amari1998natural} can be used
instead to point towards the steepest descent in the space of realizable
proposal distributions, rather than in the space of their parameters. Effectively,
the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
corrects the gradient of $U$ according to the local curvature of the KL-divergence
surface of proposal distributions, which results in forming
\begin{equation}\label{eqn:approx-grad-nat}
    \tilde{\nabla}_\psi U(\psi) = \mathbf{F}(\psi)^{-1} \nabla_\psi U(\psi),
\end{equation}
where $\mathbf{F}(\psi)$ is the Fisher information matrix
\begin{equation}
    \mathbf{F}(\psi) = \mathbb{E}_{\theta \sim q_\phi}[ \nabla_\psi \log q_\psi(\theta) \nabla_\psi \log q_\psi(\theta)^{\top} ]
\end{equation}
and can be estimated from samples, in particular by reusing the derivatives $\nabla_\psi \log q_\psi(\theta)$ already
evaluated to form $\nabla_\psi U(\psi)$.

@@

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known likelihood)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \}_{m=1}^M \sim p_r$, $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$, $\{\epsilon_m \}_{m=1}^M \sim U[0,1]$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$L^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{RMSProp}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M L^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
            \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$)}
            \State{$\psi \leftarrow \text{SGD}(\mathbf{F}(\psi)^{-1} \nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi)$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this work, we propose to rely on evolution strategies to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$g$, estimates of $\nabla_\psi U_g$ and $\tilde{\nabla}_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
% That is,
% \begin{equation}
%     \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
% \end{equation} which we can approximate with mini-batches of
% generated data
% \begin{equation}
%     \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
% \end{equation}
% for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

\glnote{Bayesian vs. frequentist view.}
Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
However, this formulation does not necessarily guarantee that the
proposal distribution $q_\psi$ will place its mass arbitrarily tight
around the parameters of interest $\theta^*$. For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] + \gamma H(q_\psi)
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
proposal distributions with larger entropy are not penalized, which may result
in learning a smeared variation of the original simulator.

For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
optimization procedure, as built on top of WGAN-GP and NES.
Let us note however that the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.
